---
title: "R Notebook"
output: html_notebook
---

## Dataset preprocessing

```{r, echo = T, include = T}
library("ggplot2")
library("survival")
library("survminer")
library(rms)

rm(list=ls())

# setwd('/home/mich/Desktop/Poli/Projects/Applied_Statistics/MDS_folder/New_Data')

MDS_data = read.csv("MDS_data.csv")
n = dim(MDS_data)[1]
```
During the preprocessing phase we

* converted all dates in date format
* converted genomic_group to character in order to use it as categorical variable
* converted genomic_group to character in order to use it as numerical variable (**maybe it's better as categorical?**)
* created dodx: first available diagnosis
* created age: age at dodx
* created time: survival time in days from first diagnosis to last follow up in days (**maybe it's better in weeks or months?**)
* created other covariates from {dodx, age, WHO}_{prev, ngs, hsct} to make the dataset more simple to use

```{r, echo = T, include = F}
# Converting date to date format
MDS_data$dodx_prev = as.Date(MDS_data$dodx_prev)
MDS_data$dodx_hsct = as.Date(MDS_data$dodx_hsct)
MDS_data$dodx_ngs = as.Date(MDS_data$dodx_ngs)
MDS_data$docyto = as.Date(MDS_data$docyto)
MDS_data$dohsct = as.Date(MDS_data$dohsct)
MDS_data$dorel = as.Date(MDS_data$dorel)
MDS_data$doaml = as.Date(MDS_data$doaml)
MDS_data$dofup = as.Date(MDS_data$dofup)

# Converting genomic_group to character
MDS_data$genomic_group <- as.character(MDS_data$genomic_group)

# Converting IPSSR to integers
MDS_data$IPSSR <- as.character(MDS_data$IPSSR)
MDS_data$IPSSR[which(MDS_data$IPSSR == "very low")] = 1
MDS_data$IPSSR[which(MDS_data$IPSSR == "low")] = 2
MDS_data$IPSSR[which(MDS_data$IPSSR == "intermediate")] = 3
MDS_data$IPSSR[which(MDS_data$IPSSR == "high")] = 4
MDS_data$IPSSR[which(MDS_data$IPSSR == "very high")] = 5
MDS_data$IPSSR = as.numeric(MDS_data$IPSSR)

# dodx: first available diagnosis
MDS_data$dodx = pmin(MDS_data$dodx_ngs, MDS_data$dodx_hsct, MDS_data$dodx_prev, na.rm = TRUE)

# date_with_age: date of first diagnosis for which we have also the age
date_with_age = pmin(MDS_data$dodx_ngs[which(!is.na(MDS_data$age_ngs))], MDS_data$dodx_hsct[which(!is.na(MDS_data$age_hsct))],  
                     MDS_data$dodx_prev[which(!is.na(MDS_data$age_prev))], na.rm = TRUE)

# age_with_date: correspondent age                                  (ASSUMPTION: there are no ages that are not associated to any diagnosis        VERIFIED!)
age_with_date = pmin(MDS_data$age_ngs, MDS_data$age_hsct, MDS_data$age_prev, na.rm = TRUE)

# age: age at dodx
MDS_data$age = age_with_date - as.integer((date_with_age - MDS_data$dodx)/365)

# time: survival time in days from first diagnosis to last follow up
MDS_data$time = MDS_data$dofup - MDS_data$dodx

# wait_time: wait time in days from first diagnosis to the transplantation
MDS_data$wait_time = MDS_data$dohsct - MDS_data$dodx


# MDS_phase: initial MDS phase ("AML" if the first diagnosis is of AML)

# if patient is DNH
#   MDS_phase = WHO_ngs, namely = MDS_{5Q- / EB1 / ...}
# if patient is HSCT
#     if WHO_hsct != AML
#       MDS_phase = WHO_hsct, namely = MDS_{5Q- / EB1 / ...}
#     else
#       if patient has _prev (we use WHO_prev since we are looking at a WHO covariate)
#         MDS_phase = WHO_prev, namely = MDS_{5Q- / EB1 / ...}
#       else
#         MDS_phase = "AML"

MDS_data$MDS_phase = NA
MDS_data$MDS_phase[which(MDS_data$cohort == "DNH")] =
  MDS_data$WHO_ngs[which(MDS_data$cohort == "DNH")]
MDS_data$MDS_phase[which(MDS_data$cohort == "HSCT" & MDS_data$WHO_hsct != "AML-MRC")] =
  MDS_data$WHO_hsct[which(MDS_data$cohort == "HSCT" & MDS_data$WHO_hsct != "AML-MRC")]
MDS_data$MDS_phase[which(MDS_data$cohort == "HSCT" & MDS_data$WHO_hsct == "AML-MRC" & !is.na(MDS_data$WHO_prev))] =
  MDS_data$WHO_prev[which(MDS_data$cohort == "HSCT" & MDS_data$WHO_hsct == "AML-MRC" & !is.na(MDS_data$WHO_prev))]
MDS_data$MDS_phase[which(MDS_data$cohort == "HSCT" & MDS_data$WHO_hsct == "AML-MRC" & is.na(MDS_data$WHO_prev))] = "AML-MRC"

# AML / do_AML: dummy variable / date associated with AML evolution 

# if patient is DNH                                            (ASSUMPTION: there are no patients in the DNH cohort with AML as first diagnosis         VERIFIED!)
#   AML_phase = AML (the original one), namely = 1 or 0
#   do_AML = doaml (the original one)
# if patient is HSCT
#     if WHO_hsct != AML
#       AML_phase = 0
#       do_AML = NA
#     else
#       AML_phase = 1                                           (ASSUMPTION: there are no AML in WHO_prev         VERIFIED!)
#       do_AML = dodx_HSCT    

MDS_data$AML_phase = MDS_data$AML
MDS_data$do_aml = MDS_data$doaml
MDS_data$AML_phase[which(MDS_data$cohort == "HSCT" & MDS_data$WHO_hsct != "AML-MRC")] = 0
MDS_data$do_aml[which(MDS_data$cohort == "HSCT" & MDS_data$WHO_hsct != "AML-MRC")] = NA
MDS_data$AML_phase[which(MDS_data$cohort == "HSCT" & MDS_data$WHO_hsct == "AML-MRC")] = 1
MDS_data$do_aml[which(MDS_data$cohort == "HSCT" & MDS_data$WHO_hsct == "AML-MRC")] =
MDS_data$dodx_hsct[which(MDS_data$cohort == "HSCT" & MDS_data$WHO_hsct == "AML-MRC")]

# age_AML: age at AML evolution
MDS_data$age_AML = MDS_data$age + as.integer((MDS_data$do_aml - MDS_data$dodx) / 365)

# MDS_data <- cbind(MDS_data, dodx, time, IPSSR)
# MDS_data <- data.frame(MDS_data, dodx, time, IPSSR)
attach(MDS_data)
```

## Kaplan-Meier and Log-rank test

We start plotting the Kaplan-Meier curve associated to the whole dataset:

```{r, echo = T, include = T}   
km_model <- survfit(Surv(time, MDS_data$fup_status)~1, data=MDS_data)
plot(km_model, 
     xlab = "Time (days)", 
     ylab = "Survival probability" 
     )                          # add "mark.time = TRUE" to see the censored data

```

Here are the curves for each genomic group

```{r, echo = T, include = T}
km_genomic = survfit(Surv(time, MDS_data$fup_status)~MDS_data$genomic_group, data=MDS_data)

ggsurvplot(km_genomic, data=MDS_data, pval = TRUE)
```

Here are the curves for each value of IPSSR:

```{r, echo = T, include = T}
km_IPSSR <- survfit(Surv(time, fup_status)~MDS_data$IPSSR, data = MDS_data)
ggsurvplot(km_IPSSR, data = MDS_data, pval = TRUE)
```

The result is consistent: higher is the IPPSR index, lower is the survival probability

We now plot the Kaplan-Meier curves between the two cohorts in every genomic group:

```{r, echo = T, include = T}
plots <- list()
km_genomic_model <- list()
km_genomic_model_diff <- list()

for (i in 0:(length(levels(as.factor(MDS_data$genomic_group)))-1)) {
  
  km_genomic_model[[i + 1]] <- survfit(Surv(time[which(MDS_data$genomic_group == i)], MDS_data$fup_status[which(MDS_data$genomic_group == i)])
                              ~MDS_data$cohort[which(MDS_data$genomic_group == i)], data = MDS_data)
  
  km_genomic_model_diff[[i + 1]] <- survdiff(Surv(time[which(MDS_data$genomic_group == i)], MDS_data$fup_status[which(MDS_data$genomic_group == i)])
           ~MDS_data$cohort[which(MDS_data$genomic_group == i)], data = MDS_data)

   #print(km_genomic_model_diff[[i + 1]])

  plots[[i + 1]] = ggsurvplot(km_genomic_model[[i + 1]],
                           data = MDS_data,
                           pval = TRUE, conf.int = FALSE,
                           risk.table = TRUE, # Add risk table
                           risk.table.col = "strata", # Change risk table color by groups
                           linetype = "strata", # Change line type by groups
                           surv.median.line = "hv", # Specify median survival
                           ggtheme = theme_bw(), # Change ggplot2 theme
                           legend.title = "Cohort",
                           legend.labs = c("DNH", "HSCT"),
                           title = paste("KM of genomic groups", toString(i), "vs cohort")
                            )
}

arrange_ggsurvplots(plots)
```
These results are hard to interpret since IPSSR is not considered and, as it's shown below, its concentration is very different between the two cohorts and,
as we noticed above, it is a significant variable.

We now plot the Kaplan-Meier curves between the two cohorts for each value of IPSSR:

```{r, echo = T, include = T}
plots <- list()
km.IPSSR.model <- list()
km.IPSSR.model.diff <- list()

for (i in 1:length(levels(as.factor(MDS_data$IPSSR)))) {
  
  km.IPSSR.model[[i]] <- survfit(Surv(time[which(MDS_data$IPSSR == i)], fup_status[which(MDS_data$IPSSR == i)])
                                    ~cohort[which(MDS_data$IPSSR == i)], data = MDS_data)
  
  km.IPSSR.model.diff[[i]] <- survdiff(Surv(time[which(MDS_data$IPSSR == i)], fup_status[which(MDS_data$IPSSR == i)])
                                          ~cohort[which(MDS_data$IPSSR == i)], data = MDS_data)

  plots[[i]] = ggsurvplot(km.IPSSR.model[[i]],
                              data = MDS_data,
                              pval = TRUE, conf.int = FALSE,
                              risk.table = TRUE, # Add risk table
                              risk.table.col = "strata", # Change risk table color by groups
                              linetype = "strata", # Change line type by groups
                              surv.median.line = "hv", # Specify median survival
                              ggtheme = theme_bw(), # Change ggplot2 theme
                              legend.title = "Cohort",
                              legend.labs = c("DNH", "HSCT"),
                              title = paste("KM of IPSSR", toString(i), "vs cohort")
  )
}

arrange_ggsurvplots(plots)
```

As we could expect, the transplantation has an overall positive effect for the most critical values of the IPSSR index.
However we cannot say the same for the first two values:
- there is no evidence to say that the curves are different for IPSSR=2 (p-value = 0.27) 
- for IPSSR = 1 the transplantation seems to be a bad idea
However these results can also be explained by the very low number of patients which such values of IPSSR in the HSCT cohort (see below). This can cause many problems
and one is the possible dominance of a specific genomic group, maybe unfavourable to the transplantation.

Here is the IPSSR distributions in the two cohorts:

```{r, echo = T, include = T}
layout(c(1,2))
plot(as.factor(MDS_data$IPSSR[which(MDS_data$cohort == "HSCT")]), main="IPSSR for HSCT")
plot(as.factor(MDS_data$IPSSR[which(MDS_data$cohort == "DNH")]), main="IPSSR for DNH")

print(paste("IPSSR for HSCT mean:", mean(MDS_data$IPSSR[which(MDS_data$cohort == "HSCT")], na.rm = TRUE)))
print(paste("IPSSR for DNH mean:", mean(MDS_data$IPSSR[which(MDS_data$cohort == "DNH")], na.rm = TRUE)))
```

Here we reproduce the previous plots but considering genomic groups 1 and 6 as a single one, as well as groups 3 and 5.
This classification is suggested in the Bersanelli's article

```{r, echo = T, include = T}
new_genomic_group = MDS_data$genomic_group
new_genomic_group[which(new_genomic_group==1)] = 16     #groups 1 and 6 toghether
new_genomic_group[which(new_genomic_group==6)] = 16     #groups 1 and 6 toghether
new_genomic_group[which(new_genomic_group==3)] = 35     #groups 3 and 5 toghether
new_genomic_group[which(new_genomic_group==5)] = 35     #groups 3 and 5 toghether
MDS_data <- cbind(MDS_data, new_genomic_group)
```

Kaplan-Meier curves for the new genomic groups

```{r, echo = T, include = T}
km.new_genomic.model <- survfit(Surv(time, MDS_data$fup_status)~MDS_data$new_genomic_group)
ggsurvplot(km.new_genomic.model,
                         data= MDS_data,
                         pval = TRUE, conf.int = FALSE,
                         risk.table = TRUE, # Add risk table
                         risk.table.col = "strata", # Change risk table color by groups
                         linetype = "strata", # Change line type by groups
                         surv.median.line = "hv", # Specify median survival
                         ggtheme = theme_bw(), # Change ggplot2 theme
                         title = "KM wrt new genomic groups")
```

This result is actually much different from the plot in the Bersanelli's article
Kaplan-Meier curves between the two cohorts in every new genomic group:

```{r, echo = T, include = T}
plots <- list()
km_new_genomic_model <- list()
km_new_genomic_model_diff <- list()
n = 1
for (i in c(0,2,4,7,16,35)) {
  km_new_genomic_model[[n]] <- survfit(Surv(time[which(MDS_data$new_genomic_group == i)], MDS_data$fup_status[which(MDS_data$new_genomic_group == i)])
                              ~MDS_data$cohort[which(MDS_data$new_genomic_group == i)], data = MDS_data)
  
  km_new_genomic_model_diff[[n]] <- survdiff(Surv(time[which(MDS_data$new_genomic_group == i)], MDS_data$fup_status[which(MDS_data$new_genomic_group == i)])
           ~MDS_data$cohort[which(MDS_data$new_genomic_group == i)], data = MDS_data)

   #print(km_new_genomic_model_diff[[n]])

  plots[[n]] = ggsurvplot(km_new_genomic_model[[n]],
                           data = MDS_data,
                           pval = TRUE, conf.int = FALSE,
                           risk.table = TRUE, # Add risk table
                           risk.table.col = "strata", # Change risk table color by groups
                           linetype = "strata", # Change line type by groups
                           surv.median.line = "hv", # Specify median survival
                           ggtheme = theme_bw(), # Change ggplot2 theme
                           legend.title = "Cohort",
                           legend.labs = c("DNH", "HSCT"),
                           title = paste("KM of new genomic groups", toString(i), "vs cohort")
                            )
  n = n+1
}

arrange_ggsurvplots(plots)
```

## Assesment of PH assumpion

Now we test the PH assumption between the two cohorts plotting the -log(-log) of the survival probablity 

```{r, echo = T, include = T}
# PH assumption through loglog test
fit_0 <-  survfit(Surv(time, MDS_data$fup_status)~MDS_data$cohort, data=MDS_data)

## Treatment variable (rx)
plot(fit_0,col=c("black", "red"), fun="cloglog")
```

The PH assumption is verified if the distance between the two functions is constant and this is not the case: this may
suggests to treat the cohort as a time dependent variable.

## Cox PH model

We start with a simple Cox model in which the only covariate is the cohort

```{r, echo = T, include = T}
cox.model = coxph(Surv(time, fup_status) ~ cohort, MDS_data)
summary(cox.model)
```

Since exp(coef(cohortHSCT)) = 1.343 it seems that a transplanted patient has hazard of 34.3% wrt a non transplanted one but it's not so simple, indeed we have to consider the fact that the transplantation is performed for some kinds of patient in conditions that we need to compare with the one of the other group.

We try to include also genomic_group

```{r, echo = T, include = T}
cox.genomic = coxph(Surv(time, fup_status) ~ cohort + genomic_group, MDS_data)
# summary(cox.genomic)
ggforest(cox.genomic, data=MDS_data)
```

Adding genomic group makes cohort seems less important (P high, coef near to zero, exp(coef) near to 1 and .95 CI which includes 0) confirming our previous hypothesis that cohort is not enough alone to infer about the transplantation effectiveness.

It seems that few genomic groups (like 1 and 6) are associated with a good prognosis (exp(coef) < 1)) while for the others the opposite. In particular for example group 2 has exp(coef) = 2.2129 and so the risk for a patient of this group is more than 2 time the risk of a patient which is not. This is in accordance with Bersanelli who says that patients in group 2 has a mutation of TP53 which is strongly associated with a bad prognosis.  

We plot the baseline

```{r, echo = T, include = T}
ggsurvplot(survfit(cox.genomic), data=MDS_data, palette = "#2E9FDF",
           censor = F,
           pval = T, conf.int = T,
           risk.table = T,
           risk.table.col = "strata",
           linetype = "strata",
           surv.median.line = "hv",
           ggtheme = theme_bw(),
           xlab = "Time (days)",
           ylab='Survival probability',
           title = "Baseline estimated survival probability"
           )
```

We plot now the adjusted survival curves for each genomic group 

```{r, echo = T, include = T}
for (i in 0:(length(levels(as.factor(genomic_group))) - 1)) {
  
  genomic_group_df <- with(MDS_data, data.frame(genomic_group = c(as.character(i)),
                                                cohort = c("HSCT", "DNH")))
  
  fit.genomic_group <- survfit(cox.genomic, newdata = genomic_group_df)
  
  plot(fit.genomic_group, conf.int=F,
       col=c("red","blue"), lwd=2, lty=1,
       xlab="Time [days]", ylab="Survival Probability",
       main = paste("Adjusted Survival Probability Plot for genomic group =", as.character((i))))
  grid()
  legend('topright', c("HSCT", "DNH"),
         lty=c(1,1,1), lwd=c(2,2,2), col=c("red","blue"))
}

```

We try with IPSSR

```{r, echo = T, include = T}
cox.IPSSR = coxph(Surv(time, fup_status) ~ cohort + IPSSR, MDS_data)
# summary(cox.IPSSR)
ggforest(cox.IPSSR, data=MDS_data)
```

Using cohort + IPSSR as covariates it seems clear what we said before, namely that it's important to consider the risk score of each patient since probably it's higher for HSCT patients than for DNH ones (as in the following plot). Now cohort seems important and HSCT patients has the half of the risk wrt DNH patients.

```{r, echo = T, include = T}
layout(c(1,2))
plot(as.factor(IPSSR[which(cohort == "HSCT")]), main="IPSSR for HSCT")
plot(as.factor(IPSSR[which(cohort == "DNH")]), main="IPSSR for DNH")

print(paste("IPSSR mean for HSCT:", mean(IPSSR[which(cohort == "HSCT")], na.rm = TRUE)))
print(paste("IPSSR mean for DNH:", mean(IPSSR[which(cohort == "DNH")], na.rm = TRUE)))

```

```{r, echo = T, include = T}
for (i in 1:(length(levels(as.factor(IPSSR))))) {
  
  IPSSR_df <- with(MDS_data, data.frame(IPSSR = c(i),
                                        cohort = c("HSCT", "DNH")))
  
  fit.IPSSR <- survfit(cox.IPSSR, newdata = IPSSR_df)
  
  plot(fit.IPSSR, conf.int=F,
       col=c("red","blue"), lwd=2, lty=1,
       xlab="Time [days]", ylab="Survival Probability",
       main = paste("Adjusted Survival Probability Plot for IPSSR =", i))
  grid()
  legend('topright', c("HSCT", "DNH"),
         lty=c(1,1,1), lwd=c(2,2,2), col=c("red","blue"))
}

```

**Maybe it's better to use IPSSR as categorical variable?**

**Maybe it's better to also include the interaction between cohort and IPSSR?**

```{r, echo = T, include = T}
MDS_data$IPSSR <- as.character(MDS_data$IPSSR)
cox.IPSSR = coxph(Surv(time, fup_status) ~ cohort + IPSSR, MDS_data)
# summary(cox.IPSSR)
ggforest(cox.IPSSR, data=MDS_data)
```
**If it's better to use IPSSR as a numerical variable, what scale we can use? We can deduce it looking at the hazard ratios in the previous plot?**

Now we try a more complete model adding both genomic group and IPSSR and also age.

```{r, echo = T, include = T}
cox.complete = coxph(Surv(time, fup_status) ~ cohort + genomic_group + IPSSR + age, MDS_data)
# summary(cox.complete)
ggforest(cox.complete, data=MDS_data)

hist(age, xlab='Age [years]', main='Histogram of age')
# summary(age)

MDS_data$agecat69 <- cut(MDS_data$age, breaks=c(0, 69, Inf), labels=c("young", "old"))
```

It seems that cohort is important and its coefficient remains similar wrt the previous model. Age seems important since it has a very low P-value but it has exp(coef) = 1.0360. So, every year, the risk of a patient increases of 3.6% which is a significant ration. To see it better, maybe we can discretize age for example in 2 groups: "young" and "old" based on its median values which is 69

```{r, echo = T, include = T}
cox.complete = coxph(Surv(time, fup_status) ~ cohort + genomic_group + IPSSR + agecat69, MDS_data)
summary(cox.complete)
ggforest(cox.complete, data=MDS_data)
```

Let's verify the GOF of our model starting from the PH assumption.

H0: PH satisfied vs H1: PH not satisfied

```{r, echo = T, include = T}
test.ph <- cox.zph(cox.complete)
test.ph 
ggcoxzph(test.ph)

# ggcoxdiagnostics(cox.complete, type = "schoenfeld")
ggcoxdiagnostics(cox.complete, type = "scaledsch") # scaled Schonefeld residuals
```

**We saw that p-values are very low so it seems that PH assumption is not satisfied. Is it true? In this case, is our Cox model anyway significant? Otherwise, can we use covariates which doesn't satisfy PH assumption as time-dependent covariates or in a Cox model stratified on them?**

**We know there are other methods to asses non-linearity and other things, is this the case to enter into the details?**

```{r, echo = T, include = T}
ggcoxdiagnostics(cox.complete, type = "martingale") # to assess nonlinearity
ggcoxdiagnostics(cox.complete, type = "deviance") # (symmetric transformation of the Martinguale residuals), to examine influential observations
```

## Cox model with time-dependent covariates

We try a Cox model with cohort as time-dependent covariate but we first need the dataset in long format.

```{r, echo = T, include = F}
df = data.frame(id, wait_time, time, fup_status, cohort, genomic_group, IPSSR)
# head(df)

# to be removed from the begging?
df = df[-which(time == 0),]
dim(df)
attach(df)

sdata <- tmerge(df, df, id=id,
                death = event(time, fup_status),
                transpl = tdc(wait_time))
df.counting <- sdata[,-(2:5)]
# head(df.counting)
```

We create a cox model with transpl as time dependent covariate (0: not transplanted at time t, 1: transplanted at time t) and we plot the baseline survival probability. In this way, we take into account the fact that, since the patient is waiting for the transplantation, he/she needs to be considered as a DNH one. 

```{r, echo = T, include = F}
cox.complete.ext = coxph(Surv(tstart, tstop, death) ~ transpl + IPSSR + genomic_group, data = df.counting)
summary(cox.complete.ext)
ggsurvplot(survfit(cox.complete.ext),
           data = df,
           ggtheme = theme_minimal())

```

**Can be this kind of Cox model a good way? Do we need to check also if we can include age at the transplantation as time dependent variable?**

## Landmark analysis

*Landmark analysis is an alternative for time dependent cox regression, isn't it? Is Landmark Analysis appropriate for our case, considering the long term effect of investigating the response of the transplantation [https://doi.org/10.1161/CIRCOUTCOMES.110.957951]? In this reference it says "Thus, for example, the landmark analysis is a recognized useful tool for removing guarantee-time bias in the cases that response to treatment occurs early after starting treatment (eg, advanced colorectal cancer), whereas in cases in which response to treatment occurs over an extended period of time (eg, chronic leukemia), use of the landmark method is not indicated.19".*
